---
title: "Overlap Report"
author: "Proteomics and Proteostasis Lab"
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: true
params:
  data:
    label: "Upload PSM file (.tsv):"
    input: file
    value: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(scales)
library(kableExtra)
```

```{r load-and-filter}
# Load raw data
raw_psm <- read.delim(
  params$data,
  header = TRUE,
  sep    = "\t"
)

n_psm <- nrow(raw_psm)

# Keep only unique PSMs and extract fraction label
psm <- raw_psm %>%
  mutate(
    Fraction = str_extract(`Spectrum.File`, "(?<=([\\\\/]))[^\\\\/.]+(?=\\.)"),
    PPM      = (`Delta.Mass` / `Calculated.Peptide.Mass`) * 1e6,
  ) %>%
  select(Fraction, Peptide, Protein, PPM)


n_fractions <- n_distinct(psm$Fraction)
```

```{r build-sets}
# One row per fraction: list of unique peptides / proteins
peptide_dict <- psm %>%
  group_by(Fraction) %>%
  summarise(
    Peptides           = list(unique(gsub(" ", "", Peptide))),
    Number_of_Peptides = n_distinct(Peptide),
    .groups = "drop"
  ) %>%
  arrange(as.numeric(str_extract(Fraction, "\\d+$")))

protein_dict <- psm %>%
  group_by(Fraction) %>%
  summarise(
    Proteins           = list(unique(Protein)),
    Number_of_Proteins = n_distinct(Protein),
    .groups = "drop"
  ) %>%
  arrange(as.numeric(str_extract(Fraction, "\\d+$")))
```

```{r jaccard-functions}
# Jaccard index between two elements of a set-list data frame
jaccard_pair <- function(dict, i, j, col = 2) {
  a     <- dict[[col]][[i]]
  b     <- dict[[col]][[j]]
  inter <- length(intersect(a, b))
  union <- length(union(a, b))
  if (union == 0) return(NA)
  (inter / union) * 100
}

# Full symmetric Jaccard matrix
build_jaccard_matrix <- function(dict) {
  n      <- nrow(dict)
  labels <- dict$Fraction
  mat    <- matrix(
    NA,
    nrow     = n,
    ncol     = n,
    dimnames = list(labels, labels)
  )
  for (i in seq_len(n)) {
    for (j in i:n) {
      score     <- jaccard_pair(dict, i, j, col = 2)
      mat[i, j] <- score
      mat[j, i] <- score
    }
  }
  mat
}

peptide_matrix <- build_jaccard_matrix(peptide_dict)
protein_matrix <- build_jaccard_matrix(protein_dict)
```

```{r heatmap-function}
plot_jaccard_heatmap <- function(
  mat,
  color_low,
  color_mid,
  color_high,
  plot_title,
  limits = NULL
) {

  fraction_order <- rownames(mat)

  mat_lower <- mat
  mat_lower[upper.tri(mat_lower)] <- NA

  df_long <- data.frame(
    fraction_row = rep(rownames(mat_lower), times = ncol(mat_lower)),
    fraction_col = rep(colnames(mat_lower), each  = nrow(mat_lower)),
    jaccard      = as.vector(mat_lower)
  )

  df_long <- df_long[!is.na(df_long$jaccard), ]
  
  df_long <- df_long %>%
    mutate(label_text = ifelse(jaccard == 100, 
                               "100", 
                               sprintf("%.2f", jaccard)))

  df_long$fraction_row <- factor(df_long$fraction_row, levels = fraction_order)
  df_long$fraction_col <- factor(df_long$fraction_col, levels = fraction_order)

  if (is.null(limits)) {
    limits <- c(0, max(df_long$jaccard, na.rm = TRUE))
  }

  values_scaled <- scales::rescale(c(limits[1], mean(limits), limits[2]))

  ggplot(df_long, aes(x = fraction_col, y = fraction_row, fill = jaccard)) +
    geom_tile(color = "white") +
    geom_text(aes(label = label_text), 
              size = 3) + 
    scale_fill_gradientn(
      colours  = c(color_low, color_mid, color_high),
      values   = values_scaled,
      limits   = limits,
      na.value = "transparent",
      name     = "Jaccard (%)"
    ) +
    labs(x = "Fraction", y = "Fraction", title = plot_title) +
    theme_classic() +
    coord_fixed() + 
    theme(
      axis.title.x = element_text(margin = margin(t = 15)),
      axis.title.y = element_text(margin = margin(r = 15)),
      axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7),
      axis.text.y  = element_text(size = 7),
      plot.title   = element_text(face = "bold", size = 11)
    )
}
```

------------------------------------------------------------------------

## Overall summary

```{r summary-table}
data.frame(
  Metric = c(
    "Total PSMs",
    "Fractions detected",
    "Unique peptides",
    "Unique proteins"
  ),
  Value = c(
    nrow(raw_psm),
    n_fractions,
    n_distinct(psm$Peptide),
    n_distinct(psm$Protein)
  )
) %>%
  kbl() %>%
  kable_paper("hover", full_width = FALSE, font_size = 17)
```

<br>

```{r barplot-peptides, fig.height=5, fig.width=8}
peptide_bar <- peptide_dict %>%
  mutate(Fraction = factor(Fraction, levels = unique(Fraction)))

y_max_pep <- max(peptide_bar$Number_of_Peptides) * 1.25

ggplot(peptide_bar, aes(x = Fraction, y = Number_of_Peptides)) +
  geom_col(alpha = 0.80, fill = "#C06C5B", color = "grey25") +
  geom_text(
    aes(label = Number_of_Peptides),
    angle = 90, vjust = 0.5, hjust = -0.15, size = 3
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, y_max_pep)) +
  labs(
    x     = "Fraction",
    y     = "Count",
    title = "Unique peptides per fraction"
  ) +
  theme_classic() +
  theme(
    axis.title.x = element_text(margin = margin(t = 15)),
    axis.title.y = element_text(margin = margin(r = 15)),
    axis.text.x  = element_text(angle = 90, vjust = 0.5),
    plot.title   = element_text(face = "bold", size = 11))
```

<br>

```{r barplot-proteins, fig.height=5, fig.width=8}
protein_bar <- protein_dict %>%
  mutate(Fraction = factor(Fraction, levels = unique(Fraction)))

y_max_pro <- max(protein_bar$Number_of_Proteins) * 1.25

ggplot(protein_bar, aes(x = Fraction, y = Number_of_Proteins)) +
  geom_col(alpha = 0.80, fill = "#457B9D", color = "grey25") +
  geom_text(
    aes(label = Number_of_Proteins),
    angle = 90, vjust = 0.5, hjust = -0.15, size = 3
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, y_max_pro)) +
  labs(
    x     = "Fraction",
    y     = "Count",
    title = "Unique proteins per fraction"
  ) +
  theme_classic() +
  theme(
    axis.title.x = element_text(margin = margin(t = 15)),
    axis.title.y = element_text(margin = margin(r = 15)),
    axis.text.x  = element_text(angle = 90, vjust = 0.5),
    plot.title   = element_text(face = "bold", size = 11))
```

------------------------------------------------------------------------

## Jaccard Index

The Jaccard index quantifies the overlap between two fractions as the percentage of shared elements out of all unique elements combined:

$$J(A,B) = \frac{|A \cap B|}{|A \cup B|} \times 100$$

A value of **0%** means the two fractions share no peptides or proteins (perfect separation), while **100%** indicates that both fractions are identical. In a well-resolved fractionation experiment, off-diagonal values are expected to be low. High overlap between adjacent fractions may suggest insufficient chromatographic resolution.

<br>

### Jaccard index — Peptides

<br>

```{r heatmap-peptides, fig.height=6, fig.width=7}
plot_jaccard_heatmap(
  mat        = peptide_matrix,
  color_low  = "#F4F1DE",
  color_mid  = "#81B29A",
  color_high = "#C06C5B",
  plot_title = "Peptide overlap between fractions (Jaccard %)",
  limits     = c(0, 10)
)
```

------------------------------------------------------------------------

### Jaccard index — Proteins

<br>

```{r heatmap-proteins, fig.height=6, fig.width=7}
plot_jaccard_heatmap(
  mat        = protein_matrix,
  color_low  = "#F4F1DE",
  color_mid  = "#6C7A89",
  color_high = "#C08497",
  plot_title = "Protein overlap between fractions (Jaccard %)",
  limits     = c(25, 50)
)
```

------------------------------------------------------------------------

## Calibration stability by fraction

Mass accuracy is expressed as the deviation between the measured and the theoretical peptide mass in parts per million (PPM):

$$\text{PPM} = \frac{\text{Delta Mass}}{\text{Calculated Peptide Mass}} \times 10^6$$

Values close to **0 PPM** indicate accurate mass calibration. A systematic shift affecting all fractions may reflect a global calibration offset, whereas fraction-specific deviations could point to instrument drift or sample-related issues during acquisition. 

<br>

### Summary
```{r ppm-stats-table}
psm %>%
  group_by(Fraction) %>%
  summarise(
    N       = n(),
    Median  = round(median(PPM, na.rm = TRUE), 3)
  ) %>%
  arrange(as.numeric(str_extract(Fraction, "\\d+$"))) %>%
  kbl(
    col.names = c("Fraction", "N", "Median (PPM)")
  ) %>%
  kable_paper("hover", full_width = FALSE, font_size = 15)
```

<br>

### Boxplot

```{r ppm-boxplot, fig.height=5, fig.width=9}
ggplot(psm, aes(x = `Fraction`, y = PPM, fill = `Fraction`)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") + 
  coord_cartesian(ylim = c(-5, 5)) + 
  theme_minimal() +
  labs(x = "Fraction",
       y = "Mass error (PPM)") +
  theme(
    axis.title.x = element_text(margin = margin(t = 15)),
    axis.title.y = element_text(margin = margin(r = 15)),
    axis.text.x = element_text(angle = 90, vjust = 0.5))
```

<hr>

<footer style="font-size: 0.75em; color: #555;">

<p><strong>For improvements, suggestions, optimizations, and inquiries, contact:</strong></p>

<p style="text-indent: 2em;">

Patricia Martínez — <a href="mailto:patricia.martinez@ispasturias.es">patricia.martinez\@ispasturias.es</a>

</p>

<p style="text-indent: 2em;">

Sergio Carrascal — <a href="mailto:sergio.carrascal@ispasturias.es">sergio.carrascal\@ispasturias.es</a>

</p>

</footer>